Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-01-24 documentation/report.tex
Almost there

In all experiments for detection of Weakly Interacting Massive Particle (WIMP) dark matter, it is essential to develop a function that can distinguish events caused by WIMP candidates from those caused by background radiation. Manually developing such a classifier is challenging, time-consuming, and necessitates detailed physical modeling.

@@

Concretely, I worked with two dark matter experiments: the PICO-60 bubble chamber \cite{pico}, and the DEAP-3600 liquid argon scintillator \cite{deap}. In both experiments, alpha particles comprise a problematic class of background radiation. In PICO-60, WIMP-like neutron calibration events are used as training data; however, alpha particles mixed into this calibration data pose a challenge to effective training. In DEAP-3600, the detector format, containing 255 spherically arranged photomultipliers, makes data processing challenging.

In both experiments, conventional classifiers are typically used. In PICO-60, it is known to be nearly optimal; however, ML was additionally used (for its automation benefits), reaching a mean of 80.2\% accuracy. In DEAP-3600, in a simulated environment, the conventional classifier removes 99.6\% of alpha background radiation, while also (undesirably) removing 91\% of simulated WIMP events.

The objective of my study was to develop novel machine learning algorithms that perform better at this task than previous methods, in the PICO-60 and DEAP-3600 dark matter experiments, and I succeeded at this task. I have lead-authored an academic paper \cite{me} on my PICO-60 research, which the PICO collaboration has reviewed and approved.

>>>>

In all experiments for detection of WIMP dark matter, it is essential to develop a classifier that can distinguish WIMP events from background radiation, in particular alpha particles. Manually developing such a classifier necessitates time-consuming physical modeling.

@@

I worked with two dark matter experiments: the PICO-60 bubble chamber \cite{pico}, and the DEAP-3600 liquid argon scintillator \cite{deap}. In PICO-60, impurities in calibration data pose an obstacle to effective training. In DEAP-3600, the detector format, containing 255 spherically arranged photomultiplier tubes (PMTs), makes data processing challenging.

In PICO-60, supervised machine learning was used, reaching a mean of 80.2\% accuracy. In DEAP-3600, in a simulated environment, a conventional classifier removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of WIMP events. My objective was to develop novel machine learning algorithms that are more accurate.

\section{Procedure}

\subsection{PICO-60}

For PICO-60, I developed and compared two sets of classification algorithms:

\begin{enumerate}
    \item I experimented further with supervised learning, exploring the following data formats to learn which was the best-performing solution:
    \begin{itemize}
        \item An 8-band Fourier transform. I applied a dense neural network with dropout \cite{dropout} and L2 regularization (which allowed for less overfitting to impure data).
        \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
        \item A raw audio waveform with a very deep 1D convolutional neural network (CNN) inspired by Dai et al \cite{verydeepconvnets}.
        \item Images captured by cameras in the detector. I trained a 2D CNN on these, to learn whether they contain any relevant information.
    \end{itemize}
    \begin{figure}[ht]
        \centering
        \subfloat{\fbox{\includegraphics[width=0.9\textwidth]{audio_cropped}}}
        \qquad
        \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
        \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
    \end{figure}
    \item To even better handle impurities, I developed two novel semi-supervised learning algorithms. The impure alpha/WIMP-like labels were removed from a large portion of the training data, and my algorithms learned more accurate labels for this data.
    \begin{itemize}
        \item Iterative Cluster Nucleation is inspired by unsupervised clustering algorithms. A neural network is trained on a labeled set, and runs inference on the unlabeled set. Unlabeled examples with predictions closest to 0 or 1 are added to the labeled set; the newly synthesized labels are more accurate than the original labels.
        \item Gravitational Differentiation uses a novel piecewise exponential function to calculate final-layer gradients. Unlabeled data ``gravitates'' to more accurate labels.
    \end{itemize}
\end{enumerate}

\subsection{DEAP-3600}

In DEAP-3600, the key challenge is the aforementioned unusual detector format: a sphere tiled with a hexagonal lattice of PMTs. I approached this problem in three different ways:

\begin{enumerate}
    \item I tried simply inputting photon counts from each PMT into a multi-layer perceptron.
    \item I developed and applied a Mercator-like cylindrical projection, and used a 2D CNN.
    \item I developed an entirely new type of CNN, called a topological CNN. Rather than a 2D image of square pixels, kernels convolve over an arbitrary topology.
\end{enumerate}

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{map_projection}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{topological}}}
    \caption{Diagrams of the map projection (left) and topological CNN (right) systems.}
\end{figure}

\section{Results}

\subsection{Statistical Practices}

All models were trained on a training set, and a separate randomly selected validation set was used to select the best-performing technique and corresponding hyperparameters (with a grid search). A test set was used to verify the performance of the model ultimately chosen.

\subsection{PICO-60}

\begin{itemize}
    \item The dense neural network, with banded Fourier transform data, increased validation accuracy to 97.3\% on validation data, up from the previous ML accuracy of 80.2\%.
    \item The full-resolution Fourier transform unexpectedly worsened accuracy to 92.2\%.
    \item The raw waveform with a 1D very deep CNN reached 94.5\% accuracy.
    \item There was no useful information in the image data; only 63.0\% accuracy was achieved.
    \item Iterative cluster nucleation, using banded FT input data, improved accuracy to 99.1\%.
    \item Gravitational differentiation was even more accurate, at 99.7\%.
    \item The best gravitational differentiation configuration was not as accurate on final test data as on validation data, but was much better than supervised learning, at 98.3\%.
\end{itemize}

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.8\textwidth]{pico_final_results}}
    \caption{Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

\subsection{DEAP-3600}

The three algorithms I developed were evaluated based on their reduction in the rate of false positives (simulated WIMP events misidentified as neck alphas) compared to the conventional 91.0\% result. Only models with at least 99.6\% neck alpha removal were considered.

\begin{itemize}
    \item The dense neural network always produced 100\% false positives.
    \item The topological CNN produced a false positive rate of 92.6\% (also worse than 91.0\%).
    \item The cylindrical projection method with a 2D CNN improved the false positive rate from 91.0\% to 75.7\% (a 63.0\% reduction)!
\end{itemize}

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.8\textwidth]{deap_final_results}}
    \caption{Alpha background and WIMP removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}

\section{Conclusion}

These results confirm that I achieved my goal. Indeed, semi-supervised learning for PICO-60 performed better than both the previous best result with machine learning, and my own supervised learning work. This significantly improves the accuracy that can be obtained quickly and without any manual optimization, allowing the team working on the experiment to iterate more quickly without the overhead of developing a conventional classifier.

The reduction in the false positive rate for DEAP-3600 provides evidence that machine learning is capable of improving the efficiency of this experiment. This may reduce the operation time required to collect sufficient data pointing towards the existence or non-existence of WIMP dark matter.

The classifiers developed during this study demonstrate great promise for machine learning in dark matter detection. Fundamentally, the problems I have solved with respect to PICO-60 and DEAP-3600 are not specific to these two experiments; they are common across many others. Thus, my algorithms are applicable in the broader field of dark matter detection.

\section{Acknowledgements}

Many thanks to Dr. Nigel Smith, Dr. Ken Clark, Dr. Carsten Krauss, Dr. Scott Fallows, Dr. Eric V\'azquez-J\'auregui, and Dr. Pierre Gorel of SNOLAB for introducing me to PICO-60 and DEAP-3600, and generously providing access to data for training.

\section*{Bibliography}

\begin{itemize}
    \item All programming for this study was done in Python 3 \cite{python}.
    \item Keras \cite{keras}, running on a TensorFlow \cite{tensorflow} backend, was used for all machine learning tasks.
    \item NumPy \cite{numpy} and SciPy \cite{scipy} were used for linear algebra and signal processing.
    \item ROOT \cite{root}, scikit-image \cite{scikit-image}, and scikit-learn \cite{scikit-learn} were used for data loading and storage.
    \item Matplotlib \cite{matplotlib} was used for data visualization.
\end{itemize}
